\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-04-26 doc/incertae_sedis.tex
Add a word.
Update.

In order to examine the results of the rooted splits-semantics above, we explore
the outcome on a series of complex cases.  These cases are intended to
illustrate some of the issues that arise when developing a semantics of
\emph{incertae sedis}, as well as the outcomes for the rooted-split semantics.
We think that they constitute one of the main contributions of the paper.  In
each case, the synthesis trees in this section are constructed using the
BUILD-based algorithm described in section \ref{sec:background}, without
decomposing them into smaller subproblems or pruning and then unpruning
taxonomy-only taxa.  Synthesis trees that are described as \emph{incertae
sedis}-aware apply the BUILD algorithm to splits constructed using equation
\ref{eq:exsetformonea}, whereas the synthesis trees that are not \emph{incertae
sedis}-aware are constructed from splits using equation
\ref{eq:exsetformonea-traditional}.

@@

When \emph{incertae sedis} clades are allowed, it is possible for an
\emph{incertae sedis} taxon to be nested within another \emph{incertae
sedis} taxon.
In such a case we must allow the higher-ranked taxon to
be placed within a sibling, while preserving the right of the lower
ranked taxon to be placed within a lower-level sibling (Figure
\ref{fig:Nested-incertae-sedis}).

@@

@@

\paragraph{Splits-based synthesis method}

Here we focus on a split-based semantics instead of a semantics based on apply
of a series of edit operations to the tree. The split-based semantics allows us
to seek trees that are simultaneously consistent with a set of split-constraints.
The use of a split-based semantics does impose some limitations. These come
primarily because in the edit-based semantics, after we place an
\emph{incertae sedis} taxon, it is no loner \emph{incertae sedis}, whereas with
the split-based semantics taxa are either always \emph{incertae sedis} or never
\emph{incertae sedis}.

@@

\subsection{Case 1}

taxonomy = (((a1,a2)A,(b1,b2)B)AB,(c1,c2)?C,D)root;

\subsubsection{If we place $c$ within $A$}

tree1 = (((a1,c1),b1),d1);

synth-with-is: ((((a1,(c1,c2)C),a2)A,(b1,b2)B)AB),(d1)D)

placement: C within A <- AB

synth-no-is: ((a1,(c1,c2)C),a1,(b1,b2)B,(d1)D)

\subsubsection{If we place $c$ within $A$ but break $C$}

tree1 = ((a1,c1),(a2,c1),b1)

synth-with-is: ((((a1,c1),(a2,c2))A,(b1,b2)B)AB, (d1)D)

synth-no-is: ??

placement:  \begin{itemize} \item c1 <- (broken) C within A <- AB.
\item c2<- (broken) C within A <- AB \end{itemize} result: we lose
$C$, but keep $A$ and $AB$.

\subsection{Case 2: nested is}

taxonomy = (((a1,a2)A,(b1,b2)B),( (
(c11,c12)C1,(c21,c22)C2,(ci1,ci2)?CI )?C) ,(d1)D)

tree1 = ((a1,(c11,(ci1,c12))),b1)

synth-with-is: places CI wihin C1, and C within A.

\subsection{Case 3: interleaving taxa}

taxonomy = ((a1,a2)A,(b1,b2)?B,(c1,c2)?c)

input tree: ((b1,c1),(b2,c2))

synth-with-is: ((a1,a2)A,((b1,c1),(b2,c2))B) or
((a1,a2)A,((b1,c1),(b2,c2))C)

Result: we name the interleaved taxon $B$ or $C$, but cannot name it
both names.

\subsection{Case 4: creation of monotypic taxa}

see test cases.
\begin{comment} \begin{lyxcode}
\textbf{Whale:}~ ?ott6145835~{[}6145835~Brandtocetus/genus{]}~<-~ott69
8424~{[}698424~Cetacea/orde r{]}~placed~under~ott4942423~{[}4942423~Ce
totherium~/~genus{]}~<-~ott698424~{[}6 98424~Cetacea/order{]}

-~Erysipelotrichaceae~broken 
\end{lyxcode} \end{comment}
Should we do this? We could say:  \begin{itemize} \item there are 331
incertae sedis taxa that are mentioned in input trees.
\begin{itemize}
\item what is the kingom / phylum / class / order / genus / species
\end{itemize} \item we placed 14 incertae sedis taxa inside a sister
taxon.
(``incertae sedis'', ``versus ``unplaced''?) \item we placed
$o$ incertae sedis taxa \emph{outside} all sister taxa.
\item we
confirmed $n$ incertae sedis taxa as being separate from all sister
(sampled) taxa?? \item we avoiding breaking 8 taxa that had IS taxa
placed inside them.
\item we allowed $z_{1}$ new taxa into the
synthesis tree that were incertae sedis.
\item we allowed $z_{2}$ new
taxa into the synthesis tree that are marked as extinct.
\item some
nodes have as many as $w$\emph{ incertae sedis} children, making them
unbrowseable when incertae sedis children are not excluded.
\item
$v_{1}$ input trees were previously excluded \emph{entirely} because
they are nested within in an incertae sedis taxon.
\item $v_{2}$ input
trees were previously excluded \emph{partially} because they are
nested within in an incertae sedis taxon.
\end{itemize} Currently the
numbers $z_{1}$ and $z_{2}$.

@@

>>>>

In order to examine the results of the rooted splits-semantics above, we explore
the outcome on a series of complex cases.  These cases are intended to
illustrate some of the issues that arise when developing an operational semantics of
\emph{incertae sedis}, as well as the outcomes for the rooted-split semantics.
We think that they constitute one of the main contributions of the paper.  In
each case, the synthesis trees in this section are constructed using the
BUILD-based algorithm described in section \ref{sec:background}, without
decomposing them into smaller subproblems or pruning and then unpruning
taxonomy-only taxa.  Synthesis trees that are described as \emph{incertae
sedis}-aware apply the BUILD algorithm to splits constructed using equation
\ref{eq:exsetformonea}, whereas the synthesis trees that are not \emph{incertae
sedis}-aware are constructed from splits using equation
\ref{eq:exsetformonea-traditional}.

@@

@@

\end{figure}
When \emph{incertae sedis} clades are allowed, it is possible for an
\emph{incertae sedis} taxon to be nested within another \emph{incertae
sedis} taxon.
In such a case we must allow the higher-ranked taxon to
be placed within a sibling, while preserving the right of the lower
ranked taxon to be placed within a lower-level sibling (Figure
\ref{fig:Nested-incertae-sedis}).

@@

\subsection{Splits-based semantics versus edit-based semantics}

Here we focus on a split-based semantics instead of a semantics based on applying
of a series of edit operations to the tree. The split-based semantics allows us
to seek trees that are simultaneously consistent with a set of split-constraints.
The use of a split-based semantics does impose some limitations. These come
primarily because in the edit-based semantics, after we place an
\emph{incertae sedis} taxon, it is no loner \emph{incertae sedis}, whereas with
the split-based semantics taxa are either always \emph{incertae sedis} or never
\emph{incertae sedis}.

@@

Should we do this? We could say:  \begin{itemize} \item there are 331
incertae sedis taxa that are mentioned in input trees.
\begin{itemize}
\item what is the kingom / phylum / class / order / genus / species
\end{itemize} \item we placed 14 incertae sedis taxa inside a sister
taxon.
(``incertae sedis'', ``versus ``unplaced''?) \item we placed
$o$ incertae sedis taxa \emph{outside} all sister taxa.
\item we
confirmed $n$ incertae sedis taxa as being separate from all sister
(sampled) taxa?? \item we avoiding breaking 8 taxa that had IS taxa
placed inside them.
\item we allowed $z_{1}$ new taxa into the
synthesis tree that were incertae sedis.
\item we allowed $z_{2}$ new
taxa into the synthesis tree that are marked as extinct.
\item some
nodes have as many as $w$\emph{ incertae sedis} children, making them
unbrowseable when incertae sedis children are not excluded.
\item
$v_{1}$ input trees were previously excluded \emph{entirely} because
they are nested within in an incertae sedis taxon.
\item $v_{2}$ input
trees were previously excluded \emph{partially} because they are
nested within in an incertae sedis taxon.
\end{itemize} Currently the
numbers $z_{1}$ and $z_{2}$.

@@

\section{Placements}
\textbf{Whale:}~ ?ott6145835~{[}6145835~Brandtocetus/genus{]}~<-~ott69
8424~{[}698424~Cetacea/orde r{]}~placed~under~ott4942423~{[}4942423~Ce
totherium~/~genus{]}~<-~ott698424~{[}6 98424~Cetacea/order{]}

-~Erysipelotrichaceae~broken
